---
title: "Kansas School District Expenditures Per Pupil"
author: "Earl F Glynn<br>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 8,
         h = 4,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r Constants}
FILE_PREFIX   <- "Expenditures-Per-Pupil"    # no spaces
```

Percentiles:  0% (min), 10%, 25%, 50% (median), 75%, 90%, 100% (max)

```{r}
quantiles <- quantiles <- c(0.0, 0.10, 0.25, 0.50, 0.75, 0.90, 1.00)
```

```{r}
SOURCE <- "Source:  KSDE Data Central - School Finance Reports"
stamp  <- paste("watchdoglab.substack.com  efg", format(Sys.time(), "%Y-%m-%d"))
```

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 10),
                axis.title            = element_text(size = 14),

                plot.caption          = element_text(hjust = c(0.0,1.0)),
                plot.caption.position = "plot",   # R 4.0.0

                plot.title.position   = "plot",
                plot.title            = element_text(size = 13))
           )

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width)#%>%      # pixels
# as.character()                             %>%      # Jupyter Only
# display_html()                                      # Jupyter Only
}
```

Redundant checks to verify the data format and content is the same across all files.

row index is zero origin

```{r}
processHeader <- function(filename)
{
  header <- read_xls(filename, n_max = 5, col_names = FALSE)
  
  SchoolYear <- str_sub(filename, 24, 32)
  
  heading    <- header[1, 1, drop = TRUE]
  
  ColumnSchoolYear <- header[4, 4, drop = TRUE]
  
  USD        <- header[5, 1, drop = TRUE]
  County     <- header[5, 3, drop = TRUE]
  FTE        <- header[5, 4, drop = TRUE]
  PerPupil   <- header[5, 5, drop = TRUE]
  
  tibble(File = filename, SchoolYear = SchoolYear,
         ColumnSchoolYear, 
         heading = heading, USD, County, FTE, PerPupil)
}
```

Assign column names here since files have school year in most columns and
we need common columns names across all files.

```{r}
processYear <- function(filename)
{
  districts <- 
    read_xls(filename, skip = 5, col_names = FALSE) %>%
    rename(
            USD              = ...1,
            `USD Name`       = ...2,
            County           = ...3,
            `FTE Enroll`     = ...4, 
            `Capital Outlay` = ...5, 
            `Bond Interest`  = ...6,
            `All Others`     = ...7,
            `Total Expend`   = ...8
          )                                         %>%
    filter(!is.na(USD))   %>%   # remove blank line before TOTALS line   
    mutate(SchoolYear = str_sub(filename, 24, 32))  %>%
    relocate(SchoolYear)
    
  districts
}
```

Create separate "wide" files for various fields in combined file

```{r}
ExtractFieldWide <- function(field) 
{
  districtsYears                               %>%
  select(SchoolYear, USD, `USD Name`, County, 
         District, {{field}})                  %>%
  pivot_wider(names_from   = SchoolYear,
              values_from  = {{field}})        %>%
  arrange(USD)  
}
```

```{r}         
SetupPlotDensity <- function(districtsYears, field, name,
                             schoolYear = "2020-2021")
{
  plotSetup <- 
    districtsYears                        %>%
    filter(startsWith(USD, "D"),                # Exclude totals
           SchoolYear == schoolYear)      %>%
  
    ggplot(aes(x = {{field}}, y = ..density..))                         + 
      geom_histogram(fill = "skyblue", colour = "grey80", bins = 20)    +
      geom_density()                                                    +
      scale_x_continuous(labels = scales::label_dollar())               + 
      labs(
            title      = paste0("Kansas School Districts - ", name,
                                " Per Pupil"),
            subtitle   = paste(schoolYear, "school year"),
            x          = paste0(name, " Per Pupil [$]"),
            y         = "Density",
            caption = c(SOURCE, stamp)
          )                             +                               
      theme(
             panel.grid.minor.x = element_blank(),
             panel.grid.minor.y = element_blank(),
             plot.caption  = element_text(hjust=c(0.0, 1.0))
           )
  
  plotSetup
}
```

# Extract from Excel Files

## Data Source 

1. Start on this page:  [Kansas Department of Education Data Central](https://datacentral.ksde.org/)

2. School Finance Reports / [School Finance Reports Warehouse](https://datacentral.ksde.org/school_finance_reports.aspx)

3. Selections

* Report format:  Excel

* Report:  Expenditures Per Pupil

*Shows the Total Expenditures per Pupil for each district, and state totals, as well as expenditures per pupil for the Capital Outlay Fund, Bond & Interest Fund and All Other Funds combined.*

Details in [Expenditures Per Pupil](https://www.ksde.org/Portals/0/School%20Finance/reports_and_publications/Expenditures/Expend%20Per%20Pupil%20Cover%2C%20State%20Totals.pdf) Report [11 pages]

* School Year:  2007-2008 through 2020-2021

* Display Report

4. Save to Excel file:  Expenditures-Per-Pupil-yyyy-yyyy.xls

## Create list of Excel files

```{r}
 files <- 
  list.files(path = ".", pattern = glob2rx("Expenditures-Per-Pupil-*.xls")) 

length(files)
```

```{r}
files
```

## Process Headers

Verify `File` names and `SchoolYear` values in header are consistent.

```{r}
headerInfo <-
  files  %>%
  map_df(processHeader)

dim(headerInfo)
```
Make sure heading is the same for all files.

```{r}
headerInfo %>% Show(height = "200px")
```

```{r}
all( (paste0(str_sub(headerInfo$SchoolYear, 1, 5), str_sub(headerInfo$SchoolYear, 8, 9)) == 
      headerInfo$ColumnSchoolYear) %>% as.vector() )
```

```{r}
all( (headerInfo$heading == "USD Expenditures Per Pupil") %>% as.vector() )
```
```{r}
all( (headerInfo$USD == "USD") %>% as.vector() )
```
```{r}
all( (headerInfo$County == "County Name") %>% as.vector() )
```
```{r}
all( (headerInfo$FTE == "FTE Enroll") %>% as.vector() )
```

```{r}
all( (headerInfo$PerPupil == "Per Pupil") %>% as.vector() )
```

With `TRUE` for all the above checks, let's assume we can assign column names to replace the ones in the files with school year in several columns.

## Process USD data 

```{r}
districtsYears <-
  files  %>%
  map_dfr(processYear)

dim(districtsYears)
```

```{r}
glimpse(districtsYears)
```

## Sample data

Recall USD Names are not unique.  `Blue Valley` school district is in both Johnson and Riley counties.

```{r}
districtsYears  %>%  head(3)  %>%  Show(bigMark = ",")
```

Let's combine `USD Name` and `County` into single `District` field, which will be unique.

```{r}
districtsYears <- 
  districtsYears    %>%
  mutate(District = paste0(`USD Name`, " [", County, "]") )
```

```{r}
districtsYears  %>%  head(3)  %>%  Show(bigMark = ",")
```

## Write combined raw data to file

### Long Format

```{r}
write_xlsx(districtsYears, 
           paste0(FILE_PREFIX, "-by-School-District-LongFormat.xlsx"))
```

## Summary stats

```{r}
districtsYears           %>%
  summarize( 
             n           = n(),
             nSchoolYear = n_distinct(SchoolYear),
             nUSD        = n_distinct(USD),
             nUSDname    = n_distinct(`USD Name`),
             nCounty     = n_distinct(County),
             nDistrict   = n_distinct(District)
           )             %>%
  Show()
```

# Full-Time Equivalent (FTE) Enrollment

## Data

```{r}
FTEWide  <- ExtractFieldWide(`FTE Enroll`)
nrow(FTEWide)
```                
Sample data

```{r}
FTEWide  %>%  head(3)  %>%  Show(caption = "FTE Enrollment")     
```

```{r}
write_xlsx(FTEWide, "FTE-Enrollment-by-School-District-WideFormat.xlsx")
```

## Density Plot - Latest Year

Shows FTE distribution across the state, which is quite broad

Override some defaults used in other plots

```{r FTE_Density}         
SetupPlotDensity(districtsYears, `FTE Enroll`,  "")     +
  scale_x_log10(breaks = c(   100,   250,   500,           # override default
                               1000,  2500,  5000, 
                              10000, 25000, 50000,
                             1000000),
                  labels = comma)                       +
    labs(title = "Kansas School Districts - Full-Time Equivalent Enrollment",
         x = "FTE Enrollment")                        
```


# Expenditures Per Pupil

## Capital Outlay

### Data

```{r}
CapitalOutlayExpendPerPupilWide  <- ExtractFieldWide(`Capital Outlay`)
nrow(CapitalOutlayExpendPerPupilWide)
```

Sample data   

```{r}
CapitalOutlayExpendPerPupilWide  %>%  head(3)  %>%  Show(caption = "Capital Outlay Expenditures Per Pupil")
```

```{r}
write_xlsx(CapitalOutlayExpendPerPupilWide,
           paste0(FILE_PREFIX, "-by-School-District-CapitalOutlay-Expenditures-WideFormat.xlsx"))
```

### Density Plot - Latest Year

```{r Capital_Outlay_Expenditures_Density} 
options(scipen = 999)
SetupPlotDensity(districtsYears, `Capital Outlay`, "Capital Outlay")  +
   scale_x_continuous(labels = scales::label_dollar(),   
                      breaks = 1000 * 0:6,
                      limits = c(0, 7000))                         
```


## Bond Interest

### Data

```{r}
BondInterestExpendPerPupilWide  <- ExtractFieldWide(`Bond Interest`)
nrow(BondInterestExpendPerPupilWide)
```

Sample data   

```{r}
BondInterestExpendPerPupilWide  %>%  head(3)  %>%  Show(caption = "Bond Interest Expenditures Per Pupil")
```

```{r}
write_xlsx(BondInterestExpendPerPupilWide,
           paste0(FILE_PREFIX, "-by-School-District-BondInterest-Expenditures-WideFormat.xlsx"))
```                                                                           

### Density Plot - Latest Year

```{r Bond_Interest_Expenditures_Density} 
SetupPlotDensity(districtsYears, `Bond Interest`, "Bond Interest")  
```

## All Others

### Data

```{r}
AllOthersExpendPerPupilWide  <- ExtractFieldWide(`All Others`)
nrow(AllOthersExpendPerPupilWide)
```

Sample data

```{r}
AllOthersExpendPerPupilWide  %>%  head(3)  %>%  Show(caption = "All Others Expenditures Per Pupil")
```

```{r}
write_xlsx(AllOthersExpendPerPupilWide,
           paste0(FILE_PREFIX, "-by-School-District-AllOthers-Expenditures-WideFormat.xlsx"))
```                                           

### Density Plot - Latest Year

```{r Other_Expenditures_Density} 
SetupPlotDensity(districtsYears, `All Others`, "All Other Expenditures")  
```


## Total Expenditures

### Data

```{r}
TotalExpendPerPupilWide  <- ExtractFieldWide(`Total Expend`)
nrow(TotalExpendPerPupilWide)
```

```{r}
TotalExpendPerPupilWide  %>%  head(3)  %>%  Show(caption = "Total Expenditures Per Pupil")
```

```{r}
write_xlsx(TotalExpendPerPupilWide,
           paste0(FILE_PREFIX, "-by-School-District-Total-Expenditures-WideFormat.xlsx"))
```

### Density Plot - Latest Year

```{r Total_Expenditures_Density}         
SetupPlotDensity(districtsYears, `Total Expend`, "Total Expenditures")  
```

# State Summary by School Year

```{r}
yearSummary <- 
  districtsYears  %>%
  filter(USD == "TOTALS") %>%
  select(SchoolYear, `FTE Enroll`, `Total Expend`)  %>%
  mutate(
          MidYear   = ymd( paste0( str_sub(SchoolYear, 6, 9), "-01-01"))
        )           %>%
  relocate(MidYear)

dim(yearSummary)
```
```{r}
yearSummary %>% Show(bigMark = ",")
```

```{r}
write_xlsx(yearSummary %>% select(-MidYear), 
           paste0(FILE_PREFIX, "-Total-Expenditures-Summary-by-School-Year.xlsx"))
```

## Governor information

```{r}
governor <-
  tibble(
          Name = c("Sebelius", "Parkinson", "Brownback",
                   "Colyer", "Kelly"),
          
          TermStart = ymd("2007-07-01",  # for plotting purposes 
                          "2009-04-28", "2011-01-10", 
                          "2018-01-31", "2019-01-14"),
          
          TermEnd   = ymd("2009-04-28", "2011-01-10", "2018-01-31",
                          "2019-01-14", 
                          "2021-12-31"),   # for plotting purposes,
          
          fraction1  = c(1.00, 0.93, 1.00, 0.93, 1.00),
          fraction2  = c(1.00, 0.60, 1.00, 0.60, 1.00),  
          
          Party      = c("D", "D", "R", "R", "D")
        )

Brownback <- governor %>% filter(Name == "Brownback")
```

## FTE State Summary by Year

```{r FTESummary}
ggplot(data = yearSummary) +
  geom_col(aes(x = MidYear, y = `FTE Enroll`),
           fill = "skyblue", colour = "grey80") +
  labs(title    = paste("FTE Enrollment of All Kansas School Districts",
                        "by School Year"),
       x = "Year",
       y = "Full-Time Equivalent Enrollment",
       caption = c(SOURCE, stamp))              +
  scale_y_continuous(labels = scales::comma,
                     breaks = 100000 * 0:5)     +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "none")               +
  geom_text(data = yearSummary,
            aes(x = MidYear, y = 0,
                label = SchoolYear),
                vjust = 0.5, hjust = -0.1, size = 4, angle = 90)   +
  geom_segment(data = governor,
               aes(x = TermStart,  y    =  fraction1 * max(yearSummary$`FTE Enroll`) * 1.1,
               xend = TermEnd,     yend =  fraction1 * max(yearSummary$`FTE Enroll`) * 1.1,
               color = Party), size = 2)        +
  geom_text(data = governor,
            aes(x = TermStart, y =  1.02 * fraction1 * max(yearSummary$`FTE Enroll`) * 1.1,
                label = Name),
                vjust = 0.0, hjust = 0, size = 5)  +
  scale_color_manual(values = c("D" = "blue", "R" = "red"))  +
  geom_vline(xintercept = c(Brownback$TermStart, Brownback$TermEnd),
             color = "red", linetype = "dashed")

 
  
```                  

## Total Expenditures per Pupil

```{r TotalExpendSummary}
ggplot(data = yearSummary) +
  geom_col(aes(x = MidYear, y = `Total Expend`),
           fill = "skyblue", colour = "grey80") +
  labs(title    = paste("Total Expenditures Per Pupil of All Kansas School Districts",
                        "by School Year"),
       x = "Year",
       y = "Total Expenditures Per Pupil [$]",
       caption = c(SOURCE, stamp))              +
   scale_y_continuous(labels = scales::label_dollar())            +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "none")               +
  geom_text(data = yearSummary,
            aes(x = MidYear, y = 0,
                label = SchoolYear),
                vjust = 0.5, hjust = -0.1, size = 4, angle = 90)  +
  geom_segment(data = governor,
               aes(x = TermStart,  y    =  fraction1 * max(yearSummary$`Total Expend`),
               xend = TermEnd,     yend =  fraction1 * max(yearSummary$`Total Expend`),
               color = Party), size = 2)        +
  geom_text(data = governor,
            aes(x = TermStart, y =  1.02 * fraction1 * max(yearSummary$`Total Expend`),
                label = Name),
                vjust = 0.0, hjust = 0, size = 5)  +
  scale_color_manual(values = c("D" = "blue", "R" = "red"))  +
  geom_vline(xintercept = c(Brownback$TermStart, Brownback$TermEnd),
             color = "red", linetype = "dashed")
```                  
  
## Total Expenditures per Pupil With Breakdown 

```{r}
yearSummaryBreakdown <- 
  districtsYears    %>%
  filter(USD == "TOTALS")                        %>%
  select(-USD, -`USD Name`, -County, -District)  %>%
  mutate(
          MidYear   = ymd( paste0( str_sub(SchoolYear, 6, 9), "-01-01"))
        )           %>%
  relocate(MidYear)

dim(yearSummaryBreakdown)
```

```{r}
yearSummaryBreakdown %>% Show(bigMark = ",")
```

```{r}
write_xlsx(yearSummaryBreakdown %>% select(-MidYear),
           paste0(FILE_PREFIX, "-Total-Expenditures-Summary-Breakdown-by-School-Year.xlsx"))
```                      


```{r}
yearSummaryBreakdownLong <- 
  yearSummaryBreakdown    %>%
  pivot_longer(
                cols      = c(`Capital Outlay`, `Bond Interest`, `All Others`),
                names_to  = "Fund",
                values_to = "Amount"
              )           %>%
  arrange(SchoolYear, Fund)
```

```{r}
PALETTE <- "Set2"   # Qualitative palette      
```

```{r TotalExpendSummaryBreakdown}
ggplot(data = yearSummaryBreakdownLong)                 +
  geom_col(aes(x = MidYear, y = Amount, fill = Fund))   +
  labs(title    = paste("Breakdown of Total Expenditures Per Pupil of All Kansas School Districts",
                        "by School Year"),
       x = "Year",
       y = "Expenditures Per Pupil [$]",
       caption = c(SOURCE, stamp))                      +
  scale_fill_brewer(palette = PALETTE)                  +      # Qualitative palette      
  scale_y_continuous(labels = scales::label_dollar())   +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "bottom")                     +
  geom_text(data = yearSummary,
            aes(x = MidYear, y = 0,
                label = SchoolYear),
                vjust = 0.5, hjust = -0.1, size = 4, angle = 90)  +
  geom_segment(data = governor,
               aes(x = TermStart,  y    =  fraction1 * max(yearSummary$`Total Expend`),
               xend = TermEnd,     yend =  fraction1 * max(yearSummary$`Total Expend`),
               color = Party), size = 2)                +
  geom_text(data = governor,
            aes(x = TermStart, y =  1.02 * fraction1 * max(yearSummary$`Total Expend`),
                label = Name),
                vjust = 0.0, hjust = 0, size = 4)       +
  scale_color_manual(values = c("D" = "blue", "R" = "red"))  +
  geom_vline(xintercept = c(Brownback$TermStart, Brownback$TermEnd),
             color = "red", linetype = "dashed")
```                  
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

