---
title: "Explore Extracting Kansas School District Cash Balances from Excel Files"
author: "Earl F Glynn<br>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
library(latex2exp)
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 10),
                axis.title            = element_text(size = 14),

                plot.caption          = element_text(hjust = c(0.0,1.0)),
                plot.caption.position = "plot",   # R 4.0.0

                plot.title.position   = "plot",
                plot.title            = element_text(size = 14),
                plot.subtitle         = element_text(size = 11)
               )
           )

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width) #  %>%
# as.character()                             %>%      # Jupyter Only
# display_html()                                      # Jupyter Only
}
```

TO DO:  Use [new way with quantiles](https://www.tidyverse.org/blog/2020/03/dplyr-1-0-0-summarise/) below

```{r}
statSummary <- function(d, variables)
{
  results <- NULL

  for (v in variables)
  {
    values <- pull(d, v)
    results <- bind_rows(results,
                         tibble(
                                Variable = v,
                                N        = length(values),
                                Sum      = sum(values, na.rm = TRUE),
                                Mean     = mean(values, na.rm = TRUE),
                                StdDev   = sd(values,   na.rm = TRUE),
                                Min      = min(values,  na.rm = TRUE),
                                Q10      = quantile(values, 0.10, na.rm = TRUE),
                                Q25      = quantile(values, 0.25, na.rm = TRUE),
                                Median   = median(values,         na.rm = TRUE),
                                Q75      = quantile(values, 0.75, na.rm = TRUE),
                                Q90      = quantile(values, 0.90, na.rm = TRUE),
                                Max      = max(values,            na.rm = TRUE)
                               )
                        )
  }

  results
}
```                          

# Data Source 

1. Start on this page:  [Kansas Department of Education Data Central](https://datacentral.ksde.org/cpfs.aspx)

2. Selections

* All Unified School Districts

* General Fund

* USD Budget Summary - Cash Balance

* School Year:  2007-2008 through 2021-2022

3. Save to Excel file:  yyyy-yyyy-Cash-Balance.xls


# Read sample file

```{r}
filename <- "2007-2008-Cash-Balance.xls"
#filename <- "2020-2021-Cash-Balance.xls"
header <- read_xls(filename, n_max = 6, col_names = FALSE)
districts <- read_xls(filename, skip = 6)
```
# Header

```{r}
header[1, 1]
```


```{r}
schoolYear <- str_sub(header[1,1], 1, 9)
schoolYear
```


```{r}
OK1 <- header[3,1] == "All Unified School Districts"
OK1
```

```{r}
OK2 <- header[5, 1] == "USD Budget Summary - Cash Balance"
OK2
```

# Data

## Raw data

```{r}
districts  %>%  head()  %>%  Show(bigMark = ",")
```

```{r}
districts  %>%  tail()  %>%  Show(bigMark = ",")
```

## Count summary

```{r}
districts                                   %>%
  summarize(n  = n(),
            nUSD = n_distinct(USD),
            nName = n_distinct(`USD Name`),
            nCounty = n_distinct(County))   %>%
  Show()
```
## Duplicate Names

```{r}
counts <- table(districts$`USD Name`)
counts[counts > 1]
```

```{r}
districts  %>%
  filter(`USD Name` == "Blue Valley")  %>%
  Show()
```


## Largest Districts

```{r}
districts                           %>%  
  slice_max(`Cash Balance`, n = 10) %>%
  Show(bigMark = ",") 
```


## Smallest Districts

```{r}
districts                           %>%  
  slice_min(`Cash Balance`, n = 10) %>%
  Show(bigMark = ",") 
```

# Density Plot

Use log scale since the range is quite large

```{r}
CAPTION <- "Source:  KSDE Data Central - Kansas Education Reports"
stamp   <- paste("watchdoglab.substack.com  efg", format(Sys.time(), "%Y-%m-%d"))
```


```{r DensityPlot}
ggplot(districts, aes(x = `Cash Balance`, y = ..density..))         + 
  geom_histogram(fill = "skyblue", colour = "grey80", bins = 20)    +
  geom_density()                                                    +
  scale_x_log10(breaks = 10^(4:8),
                labels = scales::label_dollar())                    +
  labs(title = "Kansas Unified School District Budget Summmary - Cash Balances",
       subtitle = paste("KSDE Comparative Performance and Fiscal System,", 
                        schoolYear, "school year"),
       x = "USD Cash Balance [$]",
       y = "Density",
       caption = c(CAPTION, stamp))                                 +
  theme(plot.caption  = element_text(hjust=c(0.0, 1.0)))
```


```{r}
statSummary(districts, c("Cash Balance"))  %>%  Show(bigMark = ",")
```



# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

