---
title: "Extract Kansas School District Cash Balances from Excel Files"
author: "Earl F Glynn<br>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r Constants}
FILE_PREFIX   <- "x"    # no spaces
```

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
```

```{r}
library(lubridate)
```


## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 10),
                axis.title            = element_text(size = 14),

                plot.caption          = element_text(hjust = c(0.0,1.0)),
                plot.caption.position = "plot",   # R 4.0.0

                plot.title.position   = "plot",
                title                 = element_text(size = 14))
           )

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width)#%>%      # pixels
# as.character()                             %>%      # Jupyter Only
# display_html()                                      # Jupyter Only
}
```

# Data Source 

1. Start on this page:  [Kansas Department of Education Data Central](https://datacentral.ksde.org/cpfs.aspx)

2. Selections

* All Unified School Districts

* General Fund

* USD Budget Summary - Cash Balance

* School Year:  2007-2008 through 2021-2022

3. Save to Excel file:  yyyy-yyyy-Cash-Balance.xls

Details of Cash balances can be found in [School Finance Reports](https://datacentral.ksde.org/school_finance_reports.aspx)

# Processing functions

Redundant checks to verify the data format and content is the same across all files.

```{r}
processHeader <- function(filename)
{
  header <- read_xls(filename, n_max = 6, col_names = FALSE)
  SchoolYear <- str_sub(header[1,1], 1, 9)
  Districts  <- header[3, 1, drop = TRUE] 
  Report     <- header[5, 1, drop = TRUE] 
  
  tibble(File = filename, SchoolYear = SchoolYear, Districts = Districts, Report = Report)
}
```


```{r}
processYear <- function(filename)
{
  districts <- 
    read_xls(filename, skip = 6)                 %>%
    mutate(SchoolYear = str_sub(filename, 1, 9)) %>%
    relocate(SchoolYear)
  
  districts
}
```

# Create list of Excel files

```{r}
 files <- 
  list.files(path = ".", pattern = glob2rx("*-Cash-Balance.xls")) 

length(files)
```

```{r}
files
```

# Process Headers

Verify `File` names and `SchoolYear` values in header are consistent.

```{r}
headerInfo <-
  files  %>%
  map_df(processHeader)

dim(headerInfo)
```

```{r}
headerInfo %>% Show(height = "200px")
```

```{r}
all( str_sub(headerInfo$File, 1, 9) == headerInfo$SchoolYear )
```

# Process USD data 

```{r}
districtsYears <-
  files  %>%
  map_dfr(processYear)

dim(districtsYears)
```

## Sample data

Recall USD Names are not unique.  `Blue Valley` school district is in both Johnson and Riley counties.

```{r}
districtsYears  %>%  head(3)  %>%  Show(bigMark = ",")
```

Let's combine `USD Name` and `County` into single `District` field, which will be unique.

```{r}
districtsYears <- 
  districtsYears    %>%
  mutate(District = paste0(`USD Name`, " [", County, "]") )
```

```{r}
districtsYears  %>%  head(3)  %>%  Show(bigMark = ",")
```


## Write raw data to files

### Long Format

Long format for `ggplot2` plots

```{r}
write_xlsx(districtsYears, "Cash-Balances-by-School-District-LongFormat.xlsx")
```

### Wide format

```{r}
districtsYearsWide <- 
  districtsYears      %>%
  pivot_wider(names_from   = SchoolYear,
              values_from  = `Cash Balance`)

nrow(districtsYearsWide)
```

```{r}
districtsYearsWide  %>%  head(3)  %>%  Show()
```
For humans and other programs

```{r}
write_xlsx(districtsYearsWide, "Cash-Balances-by-School-District-WideFormat.xlsx")
```

## Summary stats

```{r}
districtsYears           %>%
  summarize( 
             n           = n(),
             nSchoolYear = n_distinct(SchoolYear),
             nUSD        = n_distinct(USD),
             nUSDname    = n_distinct(`USD Name`),
             nCounty     = n_distinct(County),
             nDistrict   = n_distinct(District)
           )             %>%
  Show()
```
## Summaries by Year

```{r}
quantiles <- quantiles <- c(0.0, 0.10, 0.25, 0.50, 0.75, 0.90, 1.00)
```

### Long Format

for `ggplot2` plots

```{r}
yearSummaryLong <-
  districtsYears       %>%
  group_by(SchoolYear) %>%
  summarize(
             n         = n(),
             nDistrict = n_distinct(District),
             CashSum   = sum(`Cash Balance`, na.rm = TRUE),
             
             qValue    = quantile(`Cash Balance`, quantiles),
             Q         = quantiles,
             .groups = "drop"
           )           %>%
  ungroup()           

yearSummaryLong %>% head(14) %>% Show(bigMark = ",")
```

### Wide format

for humans and some other programs

```{r}
yearSummaryWide <- 
  yearSummaryLong  %>%
  pivot_wider(names_from   = Q,
              names_prefix = "CashQ",
              values_from  = qValue)
       
```

```{r}
yearSummaryWide %>%  Show(bigMark = ",")  
```

Negative values:

* 2010-2011:  USD 473, Chapman [Dickinson County], $-16 million  [related to Chapman tornado in 2008?]
* 2013-2014:  USD 475, Geary County Schools, $-10.6 million

```{r}
write_xlsx(yearSummaryLong, "Cash-Balance-Totals-and-Quantiles-by-School-Year-LongFormat.xlsx")
write_xlsx(yearSummaryWide, "Cash-Balance-Totals-and-Quantiles-by-School-Year-WideFormat.xlsx")
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

