---
title: "Plot Kansas School District Cash Balances -- Summary"
author: "Earl F Glynn<br>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r Constants}
FILE_PREFIX   <- "x"    # no spaces
```

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
```

```{r}
library(lubridate)
```


## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 10),
                axis.title            = element_text(size = 14),

                plot.caption          = element_text(hjust = c(0.0,1.0)),
                plot.caption.position = "plot",   # R 4.0.0

                plot.title.position   = "plot",
                plot.title            = element_text(size = 13),
                plot.subtitle         = element_text(size = 11))
           )

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width)#%>%      # pixels
# as.character()                             %>%      # Jupyter Only
# display_html()                                      # Jupyter Only
}
```

# Total Cash Balances from All School Districts

```{r}
SOURCE <- "Source:  Kansas Department of Education, School Finance Reports"
stamp  <- paste("watchdoglab.substack.com  efg", format(Sys.time(), "%Y-%m-%d"))
```

## Read summary data

```{r}
yearSummaryWide <- read_xlsx("Cash-Balance-Totals-and-Quantiles-by-School-Year-WideFormat.xlsx")
dim(yearSummaryWide)
```
By school year this table shows:

* Number of school districts

* Sum of Cash Balance for all districts

* Quantiles of Cash Balances by school district

* Q0 = 0th percentile (smallest school district cash balance)

* Q0.1 = 10th percentile

* Q0.25 = 25th percentile

* Q0.5 = 50th percentile (median school district cash balance)

* Q0.75 = 75th percentile

* Q0.9 = 90th percentile

* Q1 = 100th percentile (largest school district cash balance)



```{r}
yearSummaryWide %>% Show(bigMark = ",")
```

## Add dates for plots

The `Cash Balances` School Finance Report gives this data description:

```
Cash Balances for each fund as of July 1 and broken down by district with state totals.

*Cash balances may vary due to timing of the report or possible changes on the CPA audit that are not reflected in the USD budget data*
```

```{r}
yearSummary <- 
  yearSummaryWide   %>%
  select(SchoolYear, CashSum)  %>%
  mutate(
          MidYear   = ymd( paste0( str_sub(SchoolYear, 6, 9), "-01-01"))
        )           %>%
  relocate(MidYear)
```

```{r}
yearSummary %>%  Show(bigMark = ",")
```

## Barplot by school year

```{r}
governor <-
  tibble(
          Name = c("Sebelius", "Parkinson", "Brownback",
                   "Colyer", "Kelly"),
          
          TermStart = ymd("2007-07-01",  # for plotting purposes 
                          "2009-04-28", "2011-01-10", 
                          "2018-01-31", "2019-01-14"),
          
          TermEnd   = ymd("2009-04-28", "2011-01-10", "2018-01-31",
                          "2019-01-14", 
                          "2022-12-31"),   # for plotting purposes,
          
          fraction1  = c(1.00, 0.93, 1.00, 0.93, 1.00),
          fraction2  = c(1.00, 0.60, 1.00, 0.60, 1.00),  
          
          color     = c("D", "D", "R", "R", "D"),
          
          Q = rep(1, 5)  # Kludge for 2nd plot
        )
```

```{r}
governor %>% Show()
```
```{r}
Brownback <- governor %>% filter(Name == "Brownback")
```


```{r SummaryBarplot, fig.width = 8, fig.height = 4}
ggplot(data = yearSummary) + 
  geom_col(aes(x = MidYear, y = CashSum), 
           fill = "skyblue", colour = "grey80") +
  labs(title    = paste("Total Cash Balances of All Kansas School Districts",
                        "by School Year"),
       subtitle = "Cash Balances on July 1 (beginning of fiscal year)",
       x = "Year",
       y = "Total Cash Balances",
       caption = c(SOURCE, stamp))              +
  scale_y_continuous(labels = scales::dollar)   +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "none")               +
  geom_text(data = yearSummary,
            aes(x = MidYear, y = 20E6,
                label = SchoolYear),
                vjust = 0.5, hjust = 0, size = 4, angle = 90)  +
  geom_segment(data = governor,
               aes(x = TermStart,  y    =  fraction1 * max(yearSummary$CashSum),
               xend = TermEnd,     yend =  fraction1 * max(yearSummary$CashSum),
               linetype = "solid",
               color = color), size = 2)           +
  geom_text(data = governor,
            aes(x = TermStart, y =  1.02 * fraction1 * max(yearSummary$CashSum),
                label = Name), 
                vjust = 0.0, hjust = 0, size = 5)  +
  scale_color_manual(values = c("D" = "blue", "R" = "red"))  +
  geom_vline(xintercept = c(Brownback$TermStart, Brownback$TermEnd),
             color = "red", linetype = "dashed")
    
```
Notes:

* Total Cash Balances were the sums of Cash Balances of all Unified School Districts in Kansas.

* There were 295 school districts in 2007-2008, 293 in 2008-2009, and 286 from 2010-2011 till present.

* Values of yearly July 1 Cash Balances by school district, as well as totals of all districts, can be viewed in file Cash-Balances-by-School-District-WideFormat-Edited.xlsx.

# Variability of Cash Balance Across Kansas School Districts

```{r}
yearSummaryLong <- 
    read_xlsx("Cash-Balance-Totals-and-Quantiles-by-School-Year-LongFormat.xlsx") %>%
    select(-n, -nDistrict)  %>%
    mutate(
            StartYear   = ymd( paste0( str_sub(SchoolYear, 1, 4), "-07-01"))
          )           %>%
  relocate(StartYear)

dim(yearSummaryLong)
```

```{r}
yearSummaryLong  %>%  head(7)  %>%  Show(bigMark = ",")
```

```{r}
labels = 
  tibble(Q     = c(0, 0.1, 0.25, 0.5, 0.75, 0.9, 1),
         label = c("Smallest", "10th percentile", "25th percentile",
                   "Median", "75th percentle", "90th percentle",
                   "Largest"))


lastPoints <-
  yearSummaryLong                    %>%
  filter(SchoolYear == "2021-2022")  %>%
  inner_join(labels, by = "Q")

dim(lastPoints)
```

```{r USDQuantiles, fig.width = 8, fig.height = 4}
ggplot(yearSummaryLong, aes(x = StartYear, y = qValue, group = Q)) + 
  geom_point(size = 2, color = "black")             +
  geom_line(size = 1, color = "darkgrey", linetype = "dotted")   +
  scale_y_log10(breaks = 10^(4:8),
                labels = scales::label_dollar())    +
  scale_x_date(breaks = ymd(paste0(2007:2022, "-01-01")),
               date_labels = "%Y")   +
  labs(title    = paste("USD Cash Balances of Kansas School Districts",
                        "by School Year"),
       subtitle = "Cash Balances on July 1 (beginning of fiscal year)",
       x = "Year",
       y = "USD Cash Balances",
       caption = c(SOURCE, stamp))    + 
  theme(legend.position = "none")    +
  geom_segment(data = governor,
               aes(x = TermStart,  y    =  2 * fraction2 * max(yearSummaryLong$qValue),
               xend = TermEnd,     yend =  2 * fraction2 * max(yearSummaryLong$qValue),
               linetype = "solid",
               color = color), size = 2)      +
  geom_text(data = governor,
            aes(x = TermStart, y =  2.1 * fraction2 * max(yearSummaryLong$qValue),
                label = Name), 
                vjust = 0.0, hjust = 0, size = 5)  +
  scale_color_manual(values = c("D" = "blue", "R" = "red")) +
  geom_vline(xintercept = c(Brownback$TermStart, Brownback$TermEnd),
             color = "red", linetype = "dashed")   +
  geom_text(data = lastPoints, label = lastPoints$label,
            adj = -0.1, color = "black", size = 3)

```

Notes:  

* The "smallest" values were negative in 2010-2011 and 2013-2014 and cannot be included as part of a log scale..

* In 2010-2011 **USD 473, Chapman (Dickinson County)**, showed a deficit of $18,473,393 on 7/1/2010 for Capital Outlay (fund 16), and a Total USD Cash Balance deficit of $16,107,103.  See files Cash-Balances-2010-2011.xls and 2010-2011-Cash-Balance.xls.  

* In 2013-2014 **USD 475, Geary County Schools**, showed a deficit of 33,976,147 on 7/1/2013 for Federal (fund 07), and a Total USD Cash Balance deficit of $10,663,094.  See files Cash-Balances-2013-2014.xls and 2013-2014-Cash-Balance.xls.

* Quantile values (0.0, 0.1, 0.25, 0.5, 0.75, 0.9, 1.0) by School Year can be found in file Cash-Balance-Totals-and-Quantiles-by-School-Year-WideFormat.xlsx.  


# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

